---
title: "MLDS 491 Preliminary Analysis"
author: "Ruyi Lu"
date: "2024-11-03"
output: html_document
---

```{r}
library(tidyverse)
products = read_csv('products.csv')
countries = read_csv('countries.csv')
flavors = read_csv('flavours (redacted).csv')

length(unique(products$GNPD_PRODUCT_VARIANT_ID))
length(unique(flavors$GNPD_PRODUCT_VARIANT_ID))

data = data.frame()
data = merge(products, countries, by = "COUNTRY_CODE_2_CHAR", all.x = TRUE)
data = merge(data, flavors, by = "GNPD_PRODUCT_VARIANT_ID", all.x = TRUE)
colSums(is.na(data))

# for now, just drop all NAs for preliminary analysis
data = na.omit(data)
food = data[data$SUPERCATEGORY == 'Food',]
drink = data[data$SUPERCATEGORY == 'Drink',]

food = food %>% mutate(Month = format(as.Date(DATE_PUBLISHED), "%Y-%m"))
drink = drink %>% mutate(Month = format(as.Date(DATE_PUBLISHED), "%Y-%m"))

```

### check if month rearrangement is needed: based on eggnog(163) and pumpkin pie(1424)
```{r}
colnames(food)

eggnog <- food %>% mutate(YearMonth = floor_date(DATE_PUBLISHED, "month")) %>% filter(GNPD_FLAVOUR_ID==163) %>%
  group_by(YearMonth) %>%
  mutate(frequency = n()) %>% arrange()
eggnog$Month <- format(eggnog$YearMonth, "%m")

eggnog <- eggnog %>% group_by(Month) %>% summarise(total_freq = sum(frequency))

# Plot the total number of eggnog flavor products by month
ggplot(eggnog, aes(x = Month, y = total_freq)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Eggnog Flavor Products by Month",
       x = "Month",
       y = "Total Frequency of Eggnog Products") +
  theme_minimal()


```
So probably no need to transform the datepublished data? 

```{r}
# products
sapply(products, n_distinct)
str(products)

```


```{r}
# flavors
# sapply(flavors, n_distinct)
#str(flavors)

# food - flavor count
sapply(food, n_distinct)

# drink - flavor count
sapply(drink, n_distinct)

```

### set color for each flavor
```{r}
library(RColorBrewer)
library(viridis)

# Get 37 distinct colors (modify if you need more)
flavor_colors <- viridis::viridis_pal(option = "turbo")(37)

setequal(unique(food$FLAVOUR_CATEGORY), unique(drink$FLAVOUR_CATEGORY))

unique_flavors <- unique(food$FLAVOUR_CATEGORY)
color_palette <- setNames(flavor_colors[1:length(unique_flavors)], unique_flavors)
```


```{r}
# different types of product categories
# food
ggplot(food, aes(x = reorder(CATEGORY, CATEGORY, function(x) -length(x)), fill = CATEGORY)) +
  geom_bar() +
  labs(title = "Count of Each Product Category for Food", x = "Product Category", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8), legend.position = "bottom") +
  scale_fill_viridis_d(option = "turbo", begin = 0, end = 1, direction = 1)

# drink
ggplot(drink, aes(x = reorder(CATEGORY, CATEGORY, function(x) -length(x)), fill = CATEGORY)) +
  geom_bar() +
  labs(title = "Count of Each Product Category for Food", x = "Product Category", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8), legend.position = "bottom") +
  scale_fill_viridis_d(option = "turbo", begin = 0, end = 1, direction = 1)

```


```{r, fig.width=10, fig.height=8}
# different types of flavors
#install.packages("viridis")
library(viridis)
# food
ggplot(food, aes(x = reorder(FLAVOUR_CATEGORY, FLAVOUR_CATEGORY, function(x) -length(x)), fill = FLAVOUR_CATEGORY)) +
  geom_bar() +
  labs(title = "Count of Each Flavor Category for Food", x = "Flavor Category", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = color_palette)

# drink
ggplot(drink, aes(x = reorder(FLAVOUR_CATEGORY, FLAVOUR_CATEGORY, function(x) -length(x)), fill = FLAVOUR_CATEGORY)) +
  geom_bar() +
  labs(title = "Count of Each Flavor Category for Drink", x = "Flavor Category", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = color_palette)
```

```{r}
data %>% group_by(REGION) %>% count(NAME) %>% arrange(REGION, desc(n)) %>% ungroup()
```

```{r}
# flavors across product categories
# food
top10_flavors_f <- food %>%
  group_by(CATEGORY, FLAVOUR_CATEGORY, Month = format(as.Date(DATE_PUBLISHED), "%Y-%m")) %>%
  summarise(flavor_count = n()) %>%
  ungroup() %>%
  group_by(CATEGORY, Month) %>%
  mutate(total_count_per_month = sum(flavor_count)) %>%
  ungroup() %>%
  mutate(popularity = flavor_count / total_count_per_month) %>%
  group_by(CATEGORY, FLAVOUR_CATEGORY) %>%
  summarise(average_popularity = mean(popularity, na.rm = TRUE)) %>%
  arrange(CATEGORY, desc(average_popularity)) %>%
  slice_max(average_popularity, n = 10) %>%
  ungroup()

# drink
top10_flavors_d <- drink %>%
  group_by(CATEGORY, FLAVOUR_CATEGORY, Month = format(as.Date(DATE_PUBLISHED), "%Y-%m")) %>%
  summarise(flavor_count = n()) %>%
  ungroup() %>%
  group_by(CATEGORY, Month) %>%
  mutate(total_count_per_month = sum(flavor_count)) %>%
  ungroup() %>%
  mutate(popularity = flavor_count / total_count_per_month) %>%
  group_by(CATEGORY, FLAVOUR_CATEGORY) %>%
  summarise(average_popularity = mean(popularity, na.rm = TRUE)) %>%
  arrange(CATEGORY, desc(average_popularity)) %>%
  slice_max(average_popularity, n = 10) %>%
  ungroup()
top10_flavors_d
```

```{r, fig.width=8, fig.height=6}
#top10 flavors for each product category
# food
plot_top_flavors_f <- function(category) {
  category_data <- top10_flavors_f %>% filter(CATEGORY == category)
  ggplot(category_data, aes(x = FLAVOUR_CATEGORY, y = average_popularity, fill = FLAVOUR_CATEGORY)) +
    geom_bar(stat = "identity") +
    labs(title = paste("Food: Top 10 Flavors for", category),
         x = "Flavour Category", y = "Count") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8), legend.position = "bottom") +
    scale_fill_manual(values = color_palette)
}

lapply(unique(top10_flavors_f$CATEGORY), plot_top_flavors_f)
```


```{r, fig.width=8, fig.height=6}
#top10 flavors for each market/country name
# food
top10_flavors_f <- food %>%
  group_by(NAME, FLAVOUR_CATEGORY, Month) %>%
  summarise(flavor_count = n()) %>%
  ungroup() %>%
  group_by(NAME, Month) %>%
  mutate(total_count_per_month = sum(flavor_count)) %>%
  ungroup() %>%
  mutate(popularity = flavor_count / total_count_per_month) %>%
  group_by(NAME, FLAVOUR_CATEGORY) %>%
  summarise(average_popularity = mean(popularity, na.rm = TRUE)) %>%
  arrange(NAME, desc(average_popularity)) %>%
  slice_max(average_popularity, n = 10) %>%
  ungroup()

plot_top_flavors_f <- function(name) {
  country_data <- top10_flavors_f %>% filter(NAME == name)
  ggplot(country_data, aes(x = FLAVOUR_CATEGORY, y = average_popularity, fill = FLAVOUR_CATEGORY)) +
    geom_bar(stat = "identity") +
    labs(title = paste("Food: Top 10 Flavors for", name),
         x = "Flavour Category", y = "Count") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8), legend.position = "bottom") +
    scale_fill_manual(values = color_palette)
}

lapply(unique(top10_flavors_f$NAME), plot_top_flavors_f)
```

How to determine jump between countries: 
```{r, fig.width=8, fig.height=6}
# jump in markets
# food
popularity_by_region_f <- food %>%
  group_by(REGION, NAME, FLAVOUR_CATEGORY, Month) %>%
  summarise(flavor_count = n()) %>%
  ungroup() %>%
  group_by(NAME, Month) %>%
  mutate(total_count_per_month = sum(flavor_count)) %>%
  ungroup() %>%
  mutate(popularity = flavor_count / total_count_per_month)

popular_flavors <- popularity_by_region_f %>%
  filter(popularity > 0.1) %>%
  arrange(FLAVOUR_CATEGORY, Month, NAME)

flavor_jumps <- popular_flavors %>%
  group_by(FLAVOUR_CATEGORY) %>%
  arrange(Month) %>%
  mutate(next_market = lead(NAME),
         next_month = lead(Month)) %>%
  filter(!is.na(next_market) & next_market != NAME) %>%
  group_by(NAME, next_market) %>%
  mutate(jumps = n())

flavor_jumps

```



```{r}
#top10 flavors for each market/country name
# drink
plot_top_flavors_d <- function(category) {
  category_data <- top10_flavors_d %>% filter(CATEGORY == category)
  ggplot(category_data, aes(x = FLAVOUR_CATEGORY, y = average_popularity, fill = FLAVOUR_CATEGORY)) +
    geom_bar(stat = "identity") +
    labs(title = paste("Food: Top 10 Flavors for", category),
         x = "Flavour Category", y = "Count") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8), legend.position = "bottom") +
    scale_fill_manual(values = color_palette)
}


lapply(unique(top10_flavors_d$CATEGORY), plot_top_flavors_d)
```

```{r}
library(lubridate)
# popularity of fruit flavor across products?
flavor_selected = c("Fruit", "Chocolate", "Nut", "Berry Fruit", "Syrup", "Spice", "Citrus Fruit", "Bakery", "Vegetable", "Flower")

# food
fruit_product_over_time_f = food %>%
  mutate(Month = format(as.Date(DATE_PUBLISHED), "%Y-%m")) %>%
  filter(FLAVOUR_CATEGORY == "Fruit") %>% 
  group_by(CATEGORY, Month) %>%
  summarise(flavor_count = n(), .groups = "drop") %>%
  group_by(Month) %>%
  mutate(total_count_per_month = sum(flavor_count)) %>%
  ungroup() %>%
  mutate(popularity = flavor_count / total_count_per_month) %>%
  group_by(CATEGORY, Month) %>%
  mutate(average_popularity = mean(popularity, na.rm = TRUE)) %>%
  arrange(CATEGORY, desc(average_popularity))

print(fruit_product_over_time_f)

# drink
fruit_product_over_time_d = drink %>%
  mutate(Month = format(as.Date(DATE_PUBLISHED), "%Y-%m")) %>%
  filter(FLAVOUR_CATEGORY == "Fruit") %>% 
  group_by(CATEGORY, Month) %>%
  summarise(flavor_count = n(), .groups = "drop") %>%
  group_by(Month) %>%
  mutate(total_count_per_month = sum(flavor_count)) %>%
  ungroup() %>%
  mutate(popularity = flavor_count / total_count_per_month) %>%
  group_by(CATEGORY, Month) %>%
  mutate(average_popularity = mean(popularity, na.rm = TRUE)) %>%
  arrange(CATEGORY, desc(average_popularity))
print(fruit_product_over_time_d)
```

```{r, fig.width=10, fig.height=8}
fruit_product_over_time_f$Month <- as.Date(paste0(fruit_product_over_time_f$Month, "-01"))

ggplot(fruit_product_over_time_f, aes(x = Month, y = average_popularity, color = CATEGORY)) +
  geom_line(size = 0.75) +
  labs(title = "Food: Popularity of Fruit Flavor of Each Product Over Time",
       x = "Year & Month", y = "Average Popularity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8), legend.position = "bottom") +
  scale_color_viridis_d(option = "turbo", begin = 0, end = 1, direction = 1)

```

```{r, fig.width=10, fig.height=8}
fruit_product_over_time_d$Month <- as.Date(paste0(fruit_product_over_time_d$Month, "-01"))

ggplot(fruit_product_over_time_d, aes(x = Month, y = average_popularity, color = CATEGORY)) +
  geom_line(size = 0.75) +
  labs(title = "Drink: Popularity of Fruit Flavor of Each Product Over Time",
       x = "Year & Month", y = "Average Popularity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8), legend.position = "bottom") +
  scale_color_viridis_d(option = "turbo", begin = 0, end = 1, direction = 1)


```


```{r, fig.width=10, fig.height=8}
# Filter for Fruit flavor and calculate popularity over time for each country
fruit_popularity <- popularity_by_region_f %>%
  filter(FLAVOUR_CATEGORY == "Fruit") %>%
  group_by(REGION, NAME, Month) %>%
  summarise(average_popularity = mean(popularity, na.rm = TRUE)) %>%
  ungroup()

fruit_popularity$Month <- as.Date(paste0(fruit_popularity$Month, "-01"))

# Plot: popularity of Fruit flavor across time in each country
# Define a function to plot popularity over time for a specific region
plot_fruit_popularity_by_region <- function(region_name) {
  # Filter data for the specified region
  region_data <- popularity_by_region_f %>%
    filter(FLAVOUR_CATEGORY == "Fruit", REGION == region_name) %>%
    group_by(NAME, Month) %>%
    summarise(average_popularity = mean(popularity, na.rm = TRUE)) %>%
    ungroup()
  
  # Convert Month to Date format for accurate plotting
  region_data$Month <- as.Date(paste0(region_data$Month, "-01"))
  
  # Create the plot
  plot <- ggplot(region_data, aes(x = Month, y = average_popularity, color = NAME)) +
    geom_line(size = 0.75) +
    labs(title = paste("Popularity of Fruit Flavor Over Time in", region_name),
         x = "Time (Month)", y = "Average Popularity",
         color = "Country") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8), legend.position = "bottom")
  
  return(plot)
}

# Get unique regions (continents)
regions <- unique(popularity_by_region_f$REGION)

# Loop through each region and create a plot
plots <- list()  # To store plots
for (region in regions) {
  plot <- plot_fruit_popularity_by_region(region)
  plots[[region]] <- plot
  
  # Optionally, save each plot to a file
  ggsave(filename = paste0("fruit_popularity_", region, ".png"), plot = plot, width = 8, height = 6)
}

# Display all plots in the list (if in an interactive environment)
plots


```


Information retrieval: select features with hierarchy? how to reorganize to put things together? reasonable questions for user input? hierarchy of conditoins for data search? 
\
The type of output? the big picture question \
1. forecasting and prediction models
 - time series? launching and discontinuation
 - classification? prediction outcome? idk
 - outlier detection? abonormal event
 - future prediction?
2. visualization
 - trending plots
 - interactive dashboards
3. textual based outputs
 - key points, main patterns, general insights
 - decision making supports

\
Select features of interests? or intersted directions (textual analysis of input so too much work)
 
 
 
 
 
\
Takeaways
\
To-dos: \
1. split the dataset based on supcategory (food and drink)
2. difference between berry fruit and fruit and vegetables??? combine different flavor groups? ()
3. ***look at normalized counts instead (use prevalence or scale it? is it growing because overall amount is increasing or increasing in popularity? )
4. same product in different markets: same id?
5. look at the most important ones (top 5 in every dimension): overlappings? eg fruit flavor as important across countries? #products&region?
6. clarify what does day mean? time data entered? time product purchased? 
7. match the colors for each product or flavor? 


```{r}
#install.packages('plotly')
library(plotly)
```

```{r}
popularity_food <- food %>%
  group_by(REGION, NAME, FLAVOUR_CATEGORY, Month) %>%
  summarise(flavor_count = n()) %>%
  ungroup() %>%
  group_by(NAME, Month) %>%
  mutate(total_count_per_month = sum(flavor_count)) %>%
  ungroup() %>%
  mutate(popularity = flavor_count / total_count_per_month)
print(popularity_food)
```


```{r}
calculate_flavor_popularity <- function(data, region, country, category, create_plot = TRUE) {
  # Ensure the Month column is in the correct format
  if (!"Month" %in% colnames(data)) {
    stop("The data must contain a 'Month' column in 'YYYY-MM' format.")
  }
  
  # Check if Month is in 'YYYY-MM' format
  if (any(!grepl("^\\d{4}-\\d{2}$", data$Month))) {
    stop("The 'Month' column must be in 'YYYY-MM' format.")
  }
  
  # Filter data based on inputs
  filtered_data <- data %>%
    filter(
      REGION == region,
      NAME %in% country,  # Allow multiple countries
      CATEGORY %in% category # Allow multiple product categories
    )
  
  # Calculate popularity metrics
  popularity_food <- filtered_data %>%
    group_by(REGION, NAME, FLAVOUR_CATEGORY, CATEGORY, Month) %>%
    summarise(flavor_count = n(), .groups = "drop") %>%
    group_by(NAME, Month, CATEGORY) %>%
    mutate(total_count_per_month = sum(flavor_count)) %>%
    ungroup() %>%
    mutate(popularity = flavor_count / total_count_per_month)
  
  # Print the resulting table
  print(popularity_food)
  
  # Optional: Create a 3D plot
  if (create_plot) {
    plot <- plot_ly(
      data = popularity_food,
      x = ~NAME,                 # X-axis: Country
      y = ~CATEGORY,             # Y-axis: Product Category
      z = ~Month,                # Z-axis: Time (Month)
     # size = ~popularity,        # Dot size represents popularity
      color = ~FLAVOUR_CATEGORY, # Different colors for different flavors
      type = "scatter3d",
      mode = "markers",
      marker = list(sizemode = "diameter"),
      text = ~paste(
        "Region:", REGION,
        "<br>Country:", NAME,
        "<br>Flavor Category:", FLAVOUR_CATEGORY,
        "<br>Product Category:", CATEGORY,
        "<br>Month:", Month,
        "<br>Popularity:", flavor_count
      )
    ) %>%
      layout(
        scene = list(
          xaxis = list(title = "Country"),
          yaxis = list(title = "Product Category"),
          zaxis = list(title = "Time (Month)")
        )
      )
    
    print(plot)
  }
  
  # Return the calculated table
  return(popularity_food)
}

```


```{r}
# Example dataset
food_data <- data.frame(
  REGION = rep("Asia Pacific", 12),
  NAME = rep("Australia", 12),
  FLAVOUR_CATEGORY = rep("Alcohol", 12),
  Month = c("2015-01", "2015-02", "2015-03", "2015-04", "2015-05", "2015-06",
            "2015-07", "2015-08", "2015-09", "2015-10", "2015-11", "2015-12"),
  flavor_count = c(3, 2, 4, 1, 1, 2, 2, 2, 1, 2, 2, 4)
)

# Add total counts per month (simulate grouped data)
food_data$total_count_per_month <- c(170, 185, 200, 355, 169, 215, 226, 160, 142, 261, 261, 235)

# Add popularity
food_data$popularity <- food_data$flavor_count / food_data$total_count_per_month

```


```{r}
test_food = food %>% filter(REGION == "Europe")
test_food
typeof(c('a', 'b'))
typeof(unique(test_food$NAME))
list(unique(test_food$FLAVOUR_CATEGORY))

```

```{r}
# install.packages("viridis")
library(viridis)
```


```{r}
c = unique(test_food$NAME)
p = unique(test_food$CATEGORY)

result <- calculate_flavor_popularity(
  data = test_food,                   
  region = "Europe",               
  country = c,                
  category = p,        
  create_plot = TRUE                     
)

# View the result table
print(result)
```

